# (PART) Methods {-}

# Methods (part 1) {#methods1}

## Introduction

We describe our methods in this chapter.



## Improving `toss()`

From [chapter 2](#function), we ended up with the following `toss()` function:

```{r}
#' @title Coin toss function 
#' @description Simulates tossing a coin a given number of times
#' @param x coin object (a vector)
#' @param times number of tosses
#' @param prob vector of probabilities for each side of the coin
#' @return vector of tosses
toss <- function(x, times = 1, prob = NULL) {
  sample(x, size = times, replace = TRUE, prob = prob)
}
```

The issue with the way `toss()` has been defined so far, is that you can pass it any type of vector (not necessarily of class `"coin"`), and it will still work:

```{r}
toss(c('tic', 'tac', 'toe'))
```


To create a function `toss()` that only works for objects of class `"coin"`, we could add a `stop()` condition that checks if the argument `x` is of the right class:

```{r}
toss <- function(x, times = 1, prob = NULL) {
  if (class(x) != "coin") {
    stop("\ntoss() requires an object 'coin'")
  }
  sample(x$sides, size = times, replace = TRUE, prob = prob)
}

# ok
toss(coin1)

# bad coin
toss(c('tic', 'tac', 'toe'))
```


A more formal strategy, and one that follows OOP principles, is to create a toss __method__. In R, many functions are actually methods: e.g. `print()`, `summary()`, `plot()`, `str()`, etc. 

```{r print-method}
# print method
print
```

These types of functions are not really one unique function, they typically comprise a collection or family of functions for printing objects, computing summaries, plotting, etc. Depending on the class of the object, a generic method will look for a specific function for that class:

```{r matrix-methods}
# methods for objects "matrix"
methods(class = "matrix")
```


## Generic Method `toss`

When implementing new methods, you begin by creating a __generic__ method with the function `UseMethod()`:

```{r toss-method}
# generic method 'toss'
toss <- function(x, ...) UseMethod("toss")
```

The function `UseMethod()` allows you to declare the name of a method. In this example we are telling R that the function `toss()` is now a generic `"toss"` method. Note the use of `"..."` in the function definition, this will allow you to include more arguments when you define specific methods based on `"toss"`. 

A generic method alone is not very useful. You need to create specific cases for the generic. In our example, we only have one class `"coin"`, so that is the only class we will allow `toss` to be applied on. The way to do this is by defining `toss.coin()`:

```{r toss-coin-specific}
# specific method 'toss' for objects "coin"
toss.coin <- function(x, times = 1, prob = NULL) {
  sample(x$sides, size = times, replace = TRUE, prob = prob)
}
```

The name of the method, `"toss"`, comes first, followed by a dot `"."`, followed by the name of the class, `"coin"`. Notice that the body of the function `toss.coin()` does not include the `stop()` command anymore.

To use the `toss()` method on a `"coin"` object, you don't really have to call `toss.coin()`; calling `toss()` is enough:

```{r}
toss(coin1)
```

How does `toss()` work? Becasue `toss()` is now a generic method, everytime you use it, R will look at the class of the input, and see if there is an associated `"toss"` method. In the previous example, `coin1` is an object of class `"coin"`, for which there is a specific `toss.coin()` method. Thus using `toss()` on a `"coin"` object works fine. 

Now let's try `toss()` on the character vector `c('tic', 'tac', 'toe')`:

```{r}
# no toss() method for regular vectors
toss(c('tic', 'tac', 'toe'))
```

When you try to use `toss()` on an object that is not of class `"coin"`, you get a nice error message.



Because an object `"coin"` already contains an element `prob`, the `toss.coin()` function does not really an argument `prob`. Instead, we can pass this value from the coin object.
Here's a new definition of `toss.coin()`:

```{r new-toss}
toss.coin <- function(x, times = 1) {
  sample(x$sides, size = times, replace = TRUE, prob = x$prob)
}
```

Let's toss a loaded coin:

```{r}
set.seed(2341)
loaded_coin <- coin(c('HEADS', 'tails'), prob = c(0.75, 0.25))
toss(loaded_coin, times = 6)
```
