# Additional Methods {#methods3}

## Introduction

Until now we have standard methods like `print()`, `summary()`, and `plot()`. However, we can add more common (and not so common) methods such as:

- replacement: `"[<-.toss"`
- extraction: `"[.toss"`
- testing: `is.toss()`
- addition: `"+.toss"`


## Replacement Method

Replacement functions are those calls like `x[1] <- 3`. The function behind this expression is the replacement `"[<-"()` function. We can also create a replacement function for a given class using the notation `"[<-.class"`, where `class` is the name of the class:

```{r replacement-toss}
"[<-.toss" <- function(x, i, value) {
  if (value != x$coin$sides[1] & value != x$coin$sides[2]) {
    stop(sprintf('\nreplacing value must be %s or %s',
                 x$coin$sides[1], x$coin$sides[2]))
  }
  x$tosses[i] <- value
  make_toss(x$coin, x$tosses)
}
```

Test it:

```{r test-replacement}
set.seed(3752)
b <- toss(dime1, times = 5)
b$tosses

# replacement
b[1] <- "torch"
b$tosses
```

What about replacing out of the original range?

```{r weird-test-replacement}
# replacement in sixth position (weird!)
b[6] <- "torch"
b
```

Or something like this?

```{r fail-test-replacement}
# replacement
b[10] <- "torch"
b
```


Because it does not make sense to replace if index is out of the original length, we can add a `stop()` condition:

```{r better-replacement-toss}
"[<-.toss" <- function(x, i, value) {
  if (value != x$coin$sides[1] & value != x$coin$sides[2]) {
    stop(sprintf('\nreplacing value must be %s or %s', 
                 x$coin$sides[1], x$coin$sides[2]))
  }
  if (i > x$total) {
    stop("\nindex out of bounds")
  }
  x$tosses[i] <- value
  make_toss(x$coin, x$tosses)
}
```

Now we cannot replace if index is out of the original length:

```{r error-replacement}
set.seed(3752)
b <- toss(dime1, times = 5)
b$tosses

# replacement
b[10] <- "torch"
```


## Extraction Method

What if you want to know what is the value of toss in position 3? You could type something like this:

```{r}
b$tosses[3]
```

Or you could create an extraction method that allows you to type `x[3]`. The function behind this expression is the extraction `"["()` function. We can also create a extraction function for a given class.

```{r extract-toss}
"[.toss" <- function(x, i) {
  x$tosses[i]
}
```

Test it:

```{r test-extract}
set.seed(3752)
b <- toss(dime1, times = 5)
b$tosses
b[1]
```



## Is `"toss"`

Another common type of function for an object of a given class is `is.class()`-like functions: e.g. `is.list()`, `is.numeric()`, `is.matrix()`.
This type of functions allow you to test or check whether an object is of a given class.

We can create our own `is.toss()` to check whether a given R object is of class `"coss"`. Here's how to define such a function:

```{r}
is.toss <- function(x) {
  inherits(x, "toss")
}

is.toss(b)
is.toss(c("heads", "tails"))
```


## Addition Method

R comes with generic Math methods (see `?Math`). Among these generic methods we can find the `"+"` operator. This means that we can define our own _plus_ method for objects of class `"toss"`. The idea is to be able to call a command like this:

```{r eval = FALSE}
# toss object
b <- toss(dime1, times = 5)

# add 5 more flips
b + 5
```

Here's one implementation of `"+.toss()"` in which the first argument is an object of class `"toss"`, and the second argument is a single positive number that will play the role of additional tosses:

```{r}
"+.toss" <- function(obj, incr) {
  if (length(incr) != 1 | incr <= 0) {
    stop("\ninvalid increament (must be positive)")
  }
  more_flips <- toss(obj$coin, times = incr)
  make_toss(obj$coin, c(obj$tosses, more_flips$tosses))
}
```

Remember that `"+"` is a binary operator, which means that writing a `"+"` method requires a function with two arguments. Let's try it:

```{r}
# add four more tosses
mycoin <- coin()
seven <- toss(mycoin, times = 7)
seven
```

Let's add a couple of more tosses to `seven`:

```{r}
# two more flips
seven + 2

# three more flips
seven + 3
```
